Efficient Detection and Classification of Epigenomic Changes Under Multiple Conditions

### 2021-02-26 Latex/paper/paper.tex
txt updates

First, we segmented the genome using the Viterbi sequence of most likely HMM states to understand the distribution of genomic regions associated with consensus background, differential, and consensus enrichment states (Figure~\ref{fig:fig_multimarks}). While the majority of the genomic regions exhibited no enrichment for any of the analyzed marks, regions of consensus enrichment were rare and represented only 2\% of the analyzed genome (Figure~\ref{fig:fig_multimarks}A), as expected. Consensus background and differential regions mostly covered intergenic (66\%) and protein-coding genic regions (61\%), respectively. Differential genomic windows were mostly representing either transcribed chromatin states or repressed chromatin states (Figure~\ref{fig:fig_multimarks}B). All differential combinatorial patterns expected to be rare had associated mixture proportion estimates less than 0.05 and were pruned using our model selection scheme via BIC presented in Web Appendix B2 (Figure~\ref{fig:fig_multimarks}C). %. The Viterbi algorithm allows the simultaneous segmentation of the genome regarding the three HMM states without the necessity of choosing state-specific FDR control thresholds
\begin{figure}
\begin{center}
\includegraphics[scale = 0.7]{../../Main/Figure_MultiMarks/Figure_MultiMarks.pdf}
\end{center}
\caption{Genomic chromatin state segmentation and classification. (A): distribution of base pairs (y-axis) and the Viterbi sequence of states (x-axis). (B): estimated mixture probabilities and the associated differential combinatorial patterns. (C): choice of best model with two mixture components via BIC. (D): density estimate from expression of genes intersecting differential peaks associated with the enrichment of H3K36me3 alone or the enrichment of H3K27me3 and EZH2 in consensus. (D): association with differential H3K36me3 enrichment based based on mixture model posterior proportions of known marker genes. (E): example of a genomic region with differential peaks and genes, colored according to their classification and expression levels, respectively. This figure appears in color in the electronic version of this article.
\label{fig:fig_multimarks}}
\end{figure}

These results suggest that protein-coding genes overlapping differential regions should be either silenced (e.g. genes associated with repressed chromatin states) or highly expressed (e.g. genes associated with transcribed chromatin states). To assign combinatorial patterns to differential peaks, we chose the combination pertaining to the most frequent mixture component across windows by using the maximum estimated mixture model posterior probability, $Pr(W_{jl}=1|Z_{j}=2,\mb{y}_{..j},\mb{x};\bs{\Psi}^{(t)})$, $j=1,\ldots,M$. For genes overlapping differential regions associated with either transcribed or repressed chromatin states, we computed the distribution of transcripts per million (TPM) from matching RNA-seq experiments. Genes associated with transcribed chromatin states had a significantly higher distribution of TPM counts than genes associated with repressed chromatin states (Figure~\ref{fig:fig_multimarks}C). In fact, all known marker genes of the Helas3 cell line \citep{morin2008profiling} were associated with differential enrichment for H3K36me3, measured by the associated average mixture posterior proportions of the overlapping differential peaks (Figure~\ref{fig:fig_multimarks}D). We detected broad differential regions of enrichment and the classification of differential combinatorial patterns agreed with their biological roles as well as the expression levels of associated gene bodies. Figure~\ref{fig:fig_multimarks}E shows an example of a genomic region with differential enrichment for the analyzed marks. We compare our results to ChromHMM with 3 states, a method developed for chromatin segmentation. Our method offers the benefit of simultaneously detecting differential peaks and classifying the combinatorial pattern of enrichment through mixture model posterior probabilities even in the context of genomic segmentation with highly diverse epigenomic marks (see Web Appendix D4). By using the BIC for model selection (Web Appendix B2), one can choose the number of biologically relevant mixture components to be included in the model, a task that may not be as straightforward in methods such as ChromHMM (see Supplementary Figure 4 in \citet{ernst2012chromhmm}).

>>>>

First, we segmented the genome using the Viterbi sequence of most likely HMM states to understand the distribution of genomic regions associated with consensus background, differential, and consensus enrichment states (Figure~\ref{fig:fig_multimarks}). While the majority of the genomic regions exhibited no enrichment for any of the analyzed marks, regions of consensus enrichment were rare and represented only 2\% of the analyzed genome (Figure~\ref{fig:fig_multimarks}A), as expected. Consensus background and differential regions mostly covered intergenic (66\%) and protein-coding genic regions (61\%), respectively. Differential genomic windows were mostly representing either transcribed chromatin states or repressed chromatin states (Figure~\ref{fig:fig_multimarks}B). All differential combinatorial patterns expected to be rare had associated mixture proportion estimates less than 0.05 and were pruned using our model selection scheme via BIC presented in Web Appendix B2 (Figure~\ref{fig:fig_multimarks}C). %. The Viterbi algorithm allows the simultaneous segmentation of the genome regarding the three HMM states without the necessity of choosing state-specific FDR control thresholds
\begin{figure}
\begin{center}
\includegraphics[scale = 0.7]{../../Main/Figure_MultiMarks/Figure_MultiMarks.pdf}
\end{center}
\caption{Genomic chromatin state segmentation and classification. (A): distribution of base pairs (y-axis) and the Viterbi sequence of states (x-axis). (B): estimated mixture probabilities and the associated differential combinatorial patterns. (C): choice of best model with two mixture components via BIC. (D): density estimate from expression of genes intersecting differential peaks associated with the enrichment of H3K36me3 alone or the enrichment of H3K27me3 and EZH2 in consensus. (E): cancer marker genes and their associated mean mixture posterior proportion from differential H3K36me3 enrichment pattern of overlapping differential peaks. (F): example of a genomic region with differential peaks and genes, colored according to their classification and expression levels, respectively. This figure appears in color in the electronic version of this article.
\label{fig:fig_multimarks}}
\end{figure}

These results suggest that protein-coding genes overlapping differential regions should be either silenced (e.g. genes associated with repressed chromatin states) or highly expressed (e.g. genes associated with transcribed chromatin states). To assign combinatorial patterns to differential peaks, we chose the combination pertaining to the most frequent mixture component across windows by using the maximum estimated mixture model posterior probability, $Pr(W_{jl}=1|Z_{j}=2,\mb{y}_{..j},\mb{x};\bs{\Psi}^{(t)})$, $j=1,\ldots,M$. For genes overlapping differential regions associated with either transcribed or repressed chromatin states, we computed the distribution of transcripts per million (TPM) from matching RNA-seq experiments. Genes associated with transcribed chromatin states had a significantly higher distribution of TPM counts than genes associated with repressed chromatin states (Figure~\ref{fig:fig_multimarks}D). In fact, nearly all known Helas3 cycle-regulated genes (therefore expressed at some stage of the cell life cycle; \cite{whitfield2002identification}) were associated with differential enrichment of H3K36me3, here measured by the associated average mixture posterior proportions of the overlapping differential peaks (Figure~\ref{fig:fig_multimarks}E). The expression of such genes are known to correlate with the proliferative state of tumors and their study may help characterize their role in cancer \citep{whitfield2002identification}. We detected broad differential regions of enrichment and the classification of differential combinatorial patterns agreed with their biological roles as well as the expression levels of associated gene bodies. Figure~\ref{fig:fig_multimarks}F shows an example of a genomic region with differential enrichment for the analyzed marks. We compare our results to ChromHMM with 3 states, a method developed for chromatin segmentation. Our method offers the benefit of simultaneously detecting differential peaks and classifying the combinatorial pattern of enrichment through mixture model posterior probabilities even in the context of genomic segmentation with highly diverse epigenomic marks (see Web Appendix D4). By using the BIC for model selection (Web Appendix B2), one can choose the number of biologically relevant mixture components to be included in the model, a task that may not be as straightforward in methods such as ChromHMM (see Supplementary Figure 4 in \citet{ernst2012chromhmm}).
